NumPy---Array Programming in Python

### 2020-02-01 body.tex
Minor edits

In recentl years, there has been a rapid growth in the development of
specialized array processing hardware, including graphics processing
units (GPUs), tensor processing units (TPUs), and field-programmable
gate arrays (FPGAs).  Routinely, data sets are now larger than a
single machine's memory, and may be stored on multiple machines or in
the cloud.  Due to its in-memory data model, NumPy is unable to
utilize such specialized hardware and storage.  However, both
distributed data and the parallel execution of GPUs and TPUs map well
to the \emph{paradigm} of array programming: a gap therefore existed between
available modern hardware architectures and the tools necessary to
leverage their computational power.

@@

Such libraries often mimic the NumPy API, because it lowers the
barrier to entry for newcomers and provides the wider community with a
stable array programming interface. This, in turn, prevents disruptive
schisms like the divergence of Numeric and NumArray.
But exploring new ways of working with array is experimental by nature
and, in fact, several promising libraries---such as Caffe, Theano, and
biggus---have already ceased development. And each time that a user
decides to try a new technology or backend library, they have to
change import statements and ensure that the new library implements
all the parts of the NumPy API they currently use.

@@

  \caption{\textbf{Interoperability.} Through its array protocol, the NumPy API
     can operate on various types of array objects, including those written outside
     of NumPy. In this example, NumPy's \code{sum} function is called on a NumPy and
     a Dask array, respectively. The calls succeed by dispatching to the
     appropriate library implementation, and result in different output array
     types.
  }\label{fig:duck-arrays}

@@

% should we hint at some of the community interaction part
% they have helped us develop the neps and e.g.,
% https://github.com/pytorch/pytorch/issues/22402
To facilitate this \emph{interoperability}, NumPy provides
``protocols'' (or contracts of operation), that allow for these arrays to be
passed to NumPy functions (Fig.~\ref{fig:duck-arrays}).
NumPy, in turn, dispatches operations to the originating library, as required.
% FIXME: expand to 3 or more sentences discussing current efforts and
% issues with those, how did adoption go, what is being done next
%\url{https://numpy.org/neps/nep-0037-array-module.html}
Versions of these protocols have been successfully deployed.
% Ralf: Could you provide some proof of effectiveness, e.g., "This has made it
% possible to create distributed GPU arrays, enabling .... [find a reference to
% work by Peter Entschev]".
% Best reference: https://ep2019.europython.eu/talks/fX8dJsD-distributed-multi-gpu-computing-with-dask-cupy-and-rapids/
% Below an attempt to start on this:
Thanks to these developments users of Dask and NumPy arrays are now capable
of scaling their computation to the GPU using Dask's CuPy integration with
only minimal code changes.
As an example, consider the arrays shown in ...
\fixme{this part to be fleshed out to match the code in the figure}
% spelled out on this page: https://docs.dask.org/en/latest/gpu.html
% The video cited embedded there could be a citation:i
% https://on-demand-gtc.gputechconf.com/gtcnew/sessionview.php?sessionName=s9198-dask+and+v100s+for+fast%2c+distributed+batch+scoring+of+computer+vision+workloads
%However, general adoption has not yet been achieved mainly for compatibility
%concerns.
% explain a bit what "compatibility concerns" means ...
%Thus, we are actively addressing these concerns by revising and extending
%these protocols with the aim of wide interoperability that spans the whole
%scientific Python ecosystem up to and including domain libraries.

>>>>

In recent years, there has been a rapid growth in the development of
specialized hardware, including graphics processing
units (GPUs), tensor processing units (TPUs), and field-programmable
gate arrays (FPGAs).  Routinely, data sets are now larger than a
single machine's memory, and may be stored on multiple machines or in
the cloud.  Due to its in-memory data model, NumPy is currently unable to
utilize such specialized hardware and storage directly.  However, both
distributed data and the parallel execution of GPUs, TPUs, and FPGAs map well
to the \emph{paradigm} of array programming: a gap therefore existed between
available modern hardware architectures and the tools necessary to
leverage their computational power.

@@

Such libraries often mimic the NumPy API, because it lowers the
barrier to entry for newcomers and provides the wider community with a
stable array programming interface. This, in turn, prevents disruptive
schisms like the divergence of Numeric and NumArray.
But exploring new ways of working with arrays is experimental by nature
and, in fact, several promising libraries---such as Caffe, Theano, and
biggus---have already ceased development. And each time that a user
decides to try a new technology or backend library, they have to
change import statements and ensure that the new library implements
all the parts of the NumPy API they currently use.

@@

  \caption{\textbf{NumPy Array Protocol.} Through its array protocol, the NumPy API
     can operate on various types of array objects, including those written outside
     of NumPy. In this example, NumPy's \code{sum} function is called on a NumPy and
     a Dask array, respectively. The calls succeed by dispatching to the
     appropriate library implementation, and result in different output array
     types.
  }\label{fig:duck-arrays}

@@

To facilitate this \emph{interoperability}, NumPy provides
``protocols'' (or contracts of operation), that allow for these arrays to be
passed to NumPy functions (Fig.~\ref{fig:duck-arrays}).
NumPy, in turn, dispatches operations to the originating library, as required.
Thanks to these developments users of Dask and NumPy arrays, for instance, are now capable
of scaling their computation to the GPU using Dask's CuPy integration with
only minimal code changes.

% FIXME: expand to 3 or more sentences discussing current efforts and
% issues with those, how did adoption go, what is being done next
%\url{https://numpy.org/neps/nep-0037-array-module.html}
%Versions of these protocols have been successfully deployed.
% Ralf: Could you provide some proof of effectiveness, e.g., "This has made it
% possible to create distributed GPU arrays, enabling .... [find a reference to
% work by Peter Entschev]".
% Best reference: https://ep2019.europython.eu/talks/fX8dJsD-distributed-multi-gpu-computing-with-dask-cupy-and-rapids/
% Below an attempt to start on this:

% should we hint at some of the community interaction part
% they have helped us develop the neps and e.g.,
% https://github.com/pytorch/pytorch/issues/22402

% spelled out on this page: https://docs.dask.org/en/latest/gpu.html
% The video cited embedded there could be a citation:i
% https://on-demand-gtc.gputechconf.com/gtcnew/sessionview.php?sessionName=s9198-dask+and+v100s+for+fast%2c+distributed+batch+scoring+of+computer+vision+workloads
%However, general adoption has not yet been achieved mainly for compatibility
%concerns.
% explain a bit what "compatibility concerns" means ...
%Thus, we are actively addressing these concerns by revising and extending
%these protocols with the aim of wide interoperability that spans the whole
%scientific Python ecosystem up to and including domain libraries.
